Draft -- Discriminating Seagrasses From Green Macroalgae in European Intertidal areas using high resolution multispectral drone imagery -- Draft

### 2024-11-17 docs/_tex/index.tex
adding editing

\affiliation[1]{organization={Institut des Substances et Organismes de
la Mer, ISOMer, Nantes Université, UR 2160, F-44000 Nantes,
France},,postcodesep={}}
\affiliation[2]{organization={ECOMARE - Laboratory for Innovation and
Sustainability of Marine Biological Resources, CESAM -- Centre for
Environmental and Marine Studies, Department of Biology, University of
Aveiro, Campus Universitário de Santiago, 3810-193 Aveiro,
Portugal},,postcodesep={}}
\affiliation[3]{organization={Consiglio Nazionale delle Ricerche,
Istituto di Scienze Marine (CNR-ISMAR), 00133 Rome,
Italy},,postcodesep={}}
\affiliation[4]{organization={BRGM French Geological Survey, Cayenne
97300, French Guiana},,postcodesep={}}

@@

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
habitat fragmentation and seagrass decline. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the addition of satellite data within monitoring programs.
Yet, satellite remote sensing can be limited by too coarse spatial
and/or spectral resolution, making it difficult to discriminate seagrass
from other macrophytes in highly heterogenous meadows. Drone (unmanned
aerial vehicles -- UAV) images at a very high spatial resolution offer a
promising solution to address challenges related to spatial
heterogeneity and intrapixel mixture. This study focuses on using drone
acquisitions with a ten spectral band sensor similar to that onboard
Sentinel-2, for mapping intertidal macrophytes at low tide (i.e.~during
a period of emersion) and effectively discriminating between seagrass
and green macroalgae. Nine drone flights were conducted at two different
altitudes (12 m and 120 m) across heterogeneous intertidal European
habitats in France and Portugal, providing multispectral reflectance
observation at very high spatial resolution (8 mm and 80 mm
respectively). Taking advantage of their extremly high spatial
resolution; the low altitude flights were used to train a Neural Network
classifier to discrimintate five taxonomic classes of intertidal
vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green macroalgae),
Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae) and benthic
Bacillariophyceae (Benthic diatoms), and validated using concomitant
field measurements. Classification of drone imagery resulted in an
overall accuracy of 94\% across all sites and images, covering a total
area of 467 000 m². The model exhibited an accuracy of 96.4\% in
identifying seagrass. In particular , seagrass and green algae can be
discriminated. The very high spatial resolution of the drone data made
it possible to assess the influence of spatial resolution on the
classification outputs, showing a limited loss in seagrass detection up
to about 10m. Altogether, out findings suggest that the Multi-Spectral
Instrument (MSI) onboard Sentinel-2 offers a relevant trade-off between
its spatial and spectral resolution, thus offering promising
perspectives for satellite remote sensing of intertidal biodiversity
over lager scales.
\end{abstract}

@@

Coastal areas are vital hotspots for marine biodiversity, with
intertidal seagrass meadows playing a crucial role at the interface
between land and ocean \citep{unsworth2022}. Seagrass meadows provide a
myriad of ecosystem services, including carbon sequestration, oxygen
production, protection against sea-level rise and coastline erosion, and
mitigation of eutrophication \citep{unsworth2022, sousa2019blue}. They
serve as vital habitats for a diverse array of marine and terrestrial
species, providing living, breeding, and feeding grounds
\citep{gardner2018, Zoffoli2022, jankowska2019}. Due to the
concentration of human activities in coastal zones, seagrass meadows are
directly exposed to and impacted by anthropogenic pressures. Global
regression and fragmentation of seagrass meadows are currently observed
due to climate change, diseases, urbanization, land reclamation,
dredging, competition with alien species, and reduction in water quality
\citep{nguyen2021, soissons2018, orth2006, lin2018, duffy2019, rasheed2011long, chefaoui2018dramatic, sousa2019blue}.
Both habitat fragmentation and reduction, in turn, can severely
compromise the effectiveness of ecosystem services provided by seagrass
meadows. While improvements in water quality and hydrodynamics have been
recently reported in Europe, allowing an overall recovery of seagrass
ecosystems at local and European scales, many coastal waters worldwide
are still subjected to strong eutrophication processes
\citep{deSantos2019, Zoffoli2021, sousa2019blue}. Coastal eutrophication
has been associated to excessive accumulation of green macroalgae,
so-called green tides \citep{devlin2023nutrients}. Green tides produce
shade and suffocation over seagrass individuals, thus threatening the
health of seagrass ecosystems \citep{wang2022}.

@@

Drones (Unmanned Aerial Vehicles -- UAVs) can potentially fill the data
gaps left by satellite RS and \emph{in situ} measurements, due to their
ability to provide spatially-explicit observations at very high spatial
resolutions (pixel size from mm to cm) while capturing data at
multi-spectral resolution \citep{fairley2022drone, oh2017use}. The
versatility of drones allows for their application across a diverse
thematic range , from coastal zone management
\citep{adade2021, casella2020, angnuureng2022} to mapping species
distribution
\citep{joyce2023, tallam2023, Roca2022, Roman2021, Brunier2022Topographic, sousa2019blue}.
However, when applied to coastal habitat mapping, previous case studies
were mostly limited to a low number of drone flights over a single study
site, restricting the generalizability of their application over wider
geographical scales
\citep{Roman2021, collin2019improving, rossiter2020uav, Brunier2022Topographic}.
These studies have demonstrated the capability of drones to map
intertidal habitats, including seagrasses; however a broader
generalization of these findings is still lacking. The current paper
uniquely expands the spatial and methodological scope of drone-based
remote sensing for intertidal habitat mapping across a broad
biogeographical range. It demonstrates the feasibility of accurately
classifying diverse macrophyte types across various study sites, with a
particular focus on distinguishing Magnoliopsida (seagrasses) and
Chlorophyceae (green algae). Unlike previous studies, our approach
integrates multiple spatial scales by simulating satellite resolutions
and quantifying the impact of spatial resolution on classification
accuracy. Nine drone flights were performed over soft-bottom intertidal
areas along the Atlantic coastlines of two European countries (France
and Portugal), covering a wide range of habitats, from monospecific
seagrass meadows to meadows mixed with green, or red macroalgae. A deep
learning algorithm was trained and validated for macrophyte
discrimination, emphasizing applicability across diverse sites without
losing prediction accuracy. The classification maps obtained at a very
high spatial resolution with the drone were spatially degraded to
satellite resolutions, making it possible to assess the effect of
spatial resolution on classification accuracy, and provide insights for
coastal habitat mapping using satellite remote sensing. This study is,
therefore, among the first to quantify the effects of spatial resolution
on the accuracy of drone-based macrophyte classification across a wide
geographical scale, providing a framework to better understand
satellite-based classification challenges.

@@

Seven study sites distributed between France and Portugal were selected
for their extensive intertidal seagrass beds. Two sites were located in
the Gulf of Morbihan, France (Figure~\ref{fig-map} A : 47.5791°N,
2.8018°W). This gulf covers an area of 115 km² and is only connected to
the sea through a 900 m wide channel. A total of 53 small islands are
scattered across the gulf leading to 250 km of shorelines. Patchy
seagrass meadows can be found on many of these islands. One of the sites
within the gulf was on one its islands (Arz) and the other was located
further south on a mainland beach area (Duer). The Gulf of Morbihan is a
Natura 2000 site and a Regional Protected Area due to its rich
biodiversity, including its seagrass meadows, and is also classified as
a RAMSAR site, which highlights its significance as a wetland of
international importance. Two other sites were located in Bourgneuf Bay,
France (Figure~\ref{fig-map} B : 46.9849°N, 2.1488°W) which is a 340 km²
semi-enclosed macrotidal bay, protected from waves by Noirmoutier
Island. Bourgneuf bay hosts a large intertidal seagrass meadow of about
6 km² \citep{ZOFFOLI2020112020}. Within this meadow, the sites observed
by drones (L'Epine and Barbatre) contained monospecific beds of
\emph{Zostera noltei} (dwarf eelgrass) with very little mixing with
other macrophytes. Bourgneuf Bay is also part of the Natura 2000 network
and serves as a RAMSAR site due to its critical habitat for migratory
bird species and its extensive seagrass meadows \citep{Zoffoli2022}.
Three sites were surveyed in the Ria de Aveiro Coastal Lagoon in
Portugal (Figure~\ref{fig-map} C : 40.6887°N, 8.6810°W). The extent of
this lagoon is \textasciitilde83 km² (at low tide) with many narrow
channels, large salt marshes and many mudflats that uncover at low tide
\citep{sousa2017blue}. It is connected to the open sea through a single
channel, with a tidal lag between the North and the South of the lagoon.
The southernmost site (Gafanha) is a mudflat located in the Mira channel
(one of the four main channels of the lagoon) whereas the two other
sites (Mataducos and Marinha Lanzarote) were situated in the middle of
the lagoon and only accessible by boat. These Portuguese sites are
characterized by a more diverse intertidal vegetation, where patches of
seagrass intermingle with red, brown, and green macroalgae. The Aveiro
Lagoon, like the other study areas, is a Natura 2000 site and a RAMSAR
wetland, recognized for its rich mosaic of habitats and importance for
biodiversity, including migratory bird species and intertidal
vegetation.

@@

Before each flight, targets used as ground control points were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). Ground control points were used to
correct georeferencing imprecision of orthomosaics with an horizontal
and vertical accuracy of 10cm. A dGPS was also used to georeference
quadrats of 0.25 m², which assessed the presence or absence of five key
taxonomic classes of intertidal vegetation: Bacillariophyceae (benthic
diatoms forming biofilms at the sediment surface during low tide with
biofilm's size ranging from small patches (m²) to entire mudflats (km²);
henceforth: Benthic diatoms), Phaeophyceae (brown macroalgae generally
attached to rocks or other substrates able to form dense beds in the
intertidal zone; henceforth: Brown macroalgae), Magnoliopsida
(seagrasses, rooted flowering marine plants able to form extensive
meadows on soft sediments; henceforth: Seagrasses), Chlorophyceae (green
macroalgae, typically found attached to rocks or washed ashore;
henceforth: Green macroalgae), and Rhodophyceae (red macroalgae,
attached to hard substrates but can also be found on soft-bottom
substrate; henceforth: Red macroalgae). Only homogeneous vegetation
patches extending over several meters were selected as ground control
points. Pictures of each quadrat were uploaded online to the open-portal
Global Biodiversity Information Facility (GBIF) platform
\citep{BedeGbif}. Each photograph was also processed to estimate the
percent cover of each type of vegetation using an image processing
software \citep[ImageJ,][]{schneider2012nih}. Hyperspectral reflectance
signatures of each vegetation class were recorded using an ASD FieldSpec
HandHeld 2 spectroradiometer, which acquires reflectance between 325 and
1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures
served dual purposes: they validate the radiometric calibration of drone
data and contribute to misclassification reduction in photo
interpretations.

@@

\caption{\label{fig-vegetation}The five taxonomic classes of vegetation
used to train the Neural Network model and an example of their raw
spectral signatures at the spectral resolution of the Micasense RedEdge
Dual MX. A : Magnoliopsida (\emph{Zostera noltei}) ; B : Phaeophyceae
(\emph{Fucus sp.}) ; C : Rhodophyceae (\emph{Gracilaria
vermiculophylla}) ; D : Chlorophyceae (\emph{Ulva sp.}) ; E :
Bacillariophyceae (Benthic diatoms). The taxonomy was chacked following
the World Register of Marine Species (WORMS).}

@@

The high-altitude flight over Gafanha covered a total area of
\textasciitilde1 km² (Figure~\ref{fig-GafHigh}). A channel contouring a
small island was masked in the prediction map. Most of the vegetation
area was classified as seagrass by the model, including patches with
brown leaves. Only a few pixels were classified as green macroalgae (F1
score of 0.55). Patches of red macroalgae were correctly classified (F1
score of 0.85). In the northern part of the site and near the land
edges, patches of the schorre angiosperm \emph{Sporobolus maritimus}
(syn. \emph{Spartina maritima)} were misclassified, either as seagrass
or as brown algae (F1 score of 0.77 and 0.71, respectively).

@@

\caption{\label{fig-Boat}RGB orthomosaic (Top) and Prediction (Bottom)
of the flight made in the inner part of Ria de Aveiro Lagoon, Portugal.
The total extent of this flight was about 1.5 km² with a resolution of
80 mm per pixel. The zoom inserts cover an area equivalent to the size
of a 10-meter Sentinel-2 pixel.}

@@

The lowest performing site was Gafanha High (global accuracy of 75.45\%)
whereas Mataducos was the site with the most accurate prediction (global
accuracy of 98.05\%). Overall, the classes Phaeophyceae, Magnoliopsida,
Sediment and Rhodophyceae were correctly classified with a balanced
accuracy of 1, 0.96, 0.96 and 0.91 respectively. Bacillariophyceae was
the least accurate class (accuracy of 0.72 ) mainly due to confusion
with Magnoliopsida and Sediment.

@@

\subsection{Effect of the percent cover on the
prediction}\label{effect-of-the-percent-cover-on-the-prediction}

Using the very high-resolution low-altitude flight (8 mm pixels), we
determined the minimal percent cover required to correctly classify a
given class within the corresponding high-altitude flight (8cm pixel
resolution ; Figure~\ref{fig-upscaling}).

@@

\caption{\label{fig-upscaling}Kernel density plot showing the proportion
of pixel well classified based on the percent cover of the class in high
altitude flight pixels of Gafanha, Portugal. Each subplot shows all the
pixels of the same classes on the high altitude flight. Percent cover of
classes was retrieved using the result of the classification of the low
altitude flight of Gafanha, Portugal.}

@@

A percent cover of at least 80\% was sufficient to have all the 80 mm
pixels correctly classified, with the exception of Magnoliopsida which
required a higher cover (\textgreater90 \%) to be accurately classified.
Concerning the probability of each class, there is a linear relationship
between the percent cover and the confidence of the model to predict the
class. To predict green macroalgae with a model likelihood of 0.85, a
cover of 93 \% was needed, 90 \% for seagrass, 92 \% for red macroalgae
and 97 \% for benthic diatoms When the vegetation cover of a given class
was 100 \%, coarser high-flight pixels were correctly classified for all
the classes except for Bare Sediment, which was only correctly
classified 80\% of the time. This phenomenon may be attributed to the
time gap between the two flights, allowing for microphytobenthos
migration to the sediment surface during low tide, consequently altering
the model's classification from bare sediment to Bacillariophyceae.

@@

The authors would like to thank the extensive revisions and comments
made by anonymous reviewers to improve the work. This work was supported
through the BiCOME (Biodiversity of the Coastal Ocean: Monitoring with
Earth Observation) project funded by the European Space Agency under the
`Earth Observation Science for Society' element of FutureEO-1
BIODIVERSITY+PRECURSORS call, contract No.~4000135756/21/I-EF. This work
was also supported by French Ministry of Research \& Higher Education
for funding first authors PhD. Further, this work was partially funded
by Portuguese national funds through the FCT---Foundation for Science
and Technology, I.P., under the individual project/contract
CEECIND/00962/2017 and by FCT/MCTES through the support to CESAM
(UIDB/50017/2020+UIDP/50017/2020+LA/P/0094/2020).

>>>>

\affiliation[1]{organization={Institut des Substances et Organismes de
la Mer, ISOMer, Nantes Université, UR 2160, F-44000 Nantes,
France},,postcodesep={}}
\affiliation[2]{organization={ECOMARE, CESAM -- Centre for Environmental
and Marine Studies, Department of Biology, University of Aveiro, Campus
Universitário de Santiago, Aveiro, 3810-193, Portugal},,postcodesep={}}
\affiliation[3]{organization={Consiglio Nazionale delle Ricerche,
Istituto di Scienze Marine (CNR-ISMAR), 00133 Rome,
Italy},,postcodesep={}}
\affiliation[4]{organization={BRGM French Geological Survey, Cayenne
97300, French Guiana},,postcodesep={}}

@@

\begin{abstract}
Coastal areas support seagrass meadows, which offer crucial ecosystem
services including erosion control and carbon sequestration. However,
these areas are increasingly impacted by human activities, leading to
habitat fragmentation and seagrass decline. In situ surveys,
traditionally performed to monitor these ecosystems face limitations on
temporal and spatial coverage, particularly in intertidal zones,
prompting the addition of satellite data within monitoring programs.
Yet, satellite remote sensing can be limited by too coarse spatial
and/or spectral resolution, making it difficult to discriminate seagrass
from other macrophytes in highly heterogenous meadows. Drone (unmanned
aerial vehicles -- UAV) images at a very high spatial resolution offer a
promising solution to address challenges related to spatial
heterogeneity and intrapixel mixture. This study focuses on using drone
acquisitions with a ten spectral band sensor similar to that onboard
Sentinel-2, for mapping intertidal macrophytes at low tide (i.e.~during
a period of emersion) and effectively discriminating between seagrass
and green macroalgae. Nine drone flights were conducted at two different
altitudes (12 m and 120 m) across heterogeneous intertidal European
habitats in France and Portugal, providing multispectral reflectance
observation at very high spatial resolution (8 mm and 80 mm
respectively). Taking advantage of their extremely high spatial
resolution; the low altitude flights were used to train a Neural Network
classifier to discrimintate five taxonomic classes of intertidal
vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green macroalgae),
Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae) and benthic
Bacillariophyceae (Benthic diatoms), and validated using concomitant
field measurements. Classification of drone imagery resulted in an
overall accuracy of 94\% across all sites and images, covering a total
area of 467 000 m². The model exhibited an accuracy of 96.4\% in
identifying seagrass. In particular , seagrass and green algae can be
discriminated. The very high spatial resolution of the drone data made
it possible to assess the influence of spatial resolution on the
classification outputs, showing a limited loss in seagrass detection up
to about 10m. Altogether, our findings suggest that the Multi-Spectral
Instrument (MSI) onboard Sentinel-2 offers a relevant trade-off between
its spatial and spectral resolution, thus offering promising
perspectives for satellite remote sensing of intertidal biodiversity
over lager scales.
\end{abstract}

@@

Coastal areas are vital hotspots for marine biodiversity, with
intertidal seagrass meadows playing a crucial role at the interface
between land and ocean \citep{unsworth2022}. Seagrass meadows provide a
myriad of ecosystem services, including carbon sequestration, oxygen
production, protection against sea-level rise and coastline erosion, and
mitigation of eutrophication \citep{unsworth2022, sousa2019blue}. They
serve as vital habitats for a diverse array of marine and terrestrial
species, providing living, breeding, and feeding grounds
\citep{gardner2018, Zoffoli2022, jankowska2019}. Due to the
concentration of human activities in coastal zones, seagrass meadows are
directly exposed to and impacted by anthropogenic pressures. Global
regression and fragmentation of seagrass meadows are currently observed
due to climate change, diseases, urbanization, land reclamation,
dredging, competition with alien species, and reduction in water quality
\citep{nguyen2021, soissons2018, orth2006, lin2018, duffy2019, rasheed2011long, chefaoui2018dramatic, sousa2019blue}.
Both habitat fragmentation and reduction, in turn, can severely
compromise the effectiveness of ecosystem services provided by seagrass
meadows. While improvements in water quality and hydrodynamics have been
recently reported in Europe, allowing an overall recovery of seagrass
ecosystems at local and European scales, many coastal waters worldwide
are still subjected to strong eutrophication processes
\citep{deSantos2019, Zoffoli2021, sousa2019blue}. Coastal eutrophication
has been associated to excessive accumulation of green macroalgae (algal
blooms), so-called green tides \citep{devlin2023nutrients}. Green tides
produce shade and suffocation over seagrass individuals, thus
threatening the health of seagrass ecosystems \citep{wang2022}.

@@

Drones (Unmanned Aerial Vehicles -- UAVs) can potentially fill the data
gaps left by satellite RS and \emph{in situ} measurements, due to their
ability to provide spatially-explicit observations at very high spatial
resolutions (pixel size from mm to cm) while capturing data at
multi-spectral resolution \citep{fairley2022drone, oh2017use}. The
versatility of drones allows for their application across a diverse
thematic range , from coastal zone monitoring and management
\citep{adade2021, casella2020, angnuureng2022, lange2022large, svane2022unoccupied}
to mapping species distribution
\citep{joyce2023, tallam2023, Roca2022, Roman2021, Brunier2022Topographic, sousa2019blue, svane2022unoccupied}.
However, when applied to coastal habitat mapping, previous case studies
were mostly limited to a low number of drone flights over a single study
site, restricting the generalizability of their application over wider
geographical scales
\citep{Roman2021, collin2019improving, rossiter2020uav, Brunier2022Topographic}.
These studies have demonstrated the capability of drones to map
intertidal (and subtidal) habitats, including seagrasses; however a
broader generalization of these findings is still lacking. The current
paper uniquely expands the spatial and methodological scope of
drone-based remote sensing for intertidal habitat mapping across a broad
biogeographical range. It demonstrates the feasibility of accurately
classifying diverse macrophyte types across various study sites, with a
particular focus on distinguishing Magnoliopsida (seagrasses) and
Chlorophyceae (green algae). Unlike previous studies, our approach
integrates multiple spatial scales by simulating satellite resolutions
and quantifying the impact of spatial resolution on classification
accuracy. Nine drone flights were performed over soft-bottom (mud and
sand) intertidal areas along the Atlantic coastlines of two European
countries (France and Portugal), covering a wide range of habitats, from
monospecific seagrass meadows to meadows mixed with green, or red
macroalgae. A deep learning algorithm was trained and validated for
macrophyte discrimination, emphasizing applicability across diverse
sites without losing prediction accuracy. The classification maps
obtained at a very high spatial resolution with the drone were spatially
degraded to satellite resolutions, making it possible to assess the
effect of spatial resolution on classification accuracy, and provide
insights for coastal habitat mapping using satellite remote sensing.
This study is, therefore, among the first to quantify the effects of
spatial resolution on the accuracy of drone-based intertidal macrophyte
classification across a wide geographical scale, providing a framework
to better understand satellite-based classification challenges.

@@

Seven study sites distributed between France and Portugal were selected
for their extensive intertidal seagrass beds. Two sites were located in
the Gulf of Morbihan, France (Figure~\ref{fig-map} A : 47.5791°N,
2.8018°W). This gulf covers an area of 115 km² and is only connected to
the sea through a 900 m wide channel. A total of 53 small islands are
scattered across the gulf leading to 250 km of shorelines. Patchy
seagrass meadows can be found on many of these islands. One of the sites
within the gulf was on one its islands (Arz) and the other was located
further south on a mainland beach area (Duer). The Gulf of Morbihan is a
Natura 2000 site and a Regional Protected Area due to its rich
biodiversity, including its seagrass meadows, and is also classified as
a RAMSAR site, which highlights its significance as a wetland of
international importance. Two other sites were located in Bourgneuf Bay,
France (Figure~\ref{fig-map} B : 46.9849°N, 2.1488°W) which is a 340 km²
semi-enclosed macrotidal bay, protected from waves by Noirmoutier
Island. Bourgneuf bay hosts a large intertidal seagrass meadow of about
6 km² \citep{ZOFFOLI2020112020}. Within this meadow, the sites observed
by drones (L'Epine and Barbatre) contained monospecific beds of
\emph{Zostera noltei} (dwarf eelgrass) with very little mixing with
other macrophytes. Bourgneuf Bay is also part of the Natura 2000 network
and serves as a RAMSAR site due to its critical habitat for migratory
bird species and its extensive seagrass meadows \citep{Zoffoli2022}.
Three sites were surveyed in the Ria de Aveiro Coastal Lagoon in
Portugal (Figure~\ref{fig-map} C : 40.6887°N, 8.6810°W). The extent of
this lagoon is \textasciitilde83 km² (at low tide) with many narrow
channels, large salt marshes and many mudflats that uncover at low tide
\citep{sousa2017blue}. It is connected to the open sea through a single
channel, with a tidal lag between the North and the South of the lagoon.
The southernmost site (Gafanha) is a mudflat located in the Mira channel
(one of the four main channels of the lagoon) whereas the two other
sites (Mataduços and Marinha Lanzarote) were situated in the middle of
the lagoon, with Marinha Lanzarote only accessible by boat. These
Portuguese sites (at Ria de Aveiro) are characterized by a more diverse
intertidal vegetation, where patches of seagrass intermingle with red,
brown, and green macroalgae. The Ria de Aveiro costal lagoon, like the
other study areas, is a Natura 2000 site, recognized for its rich mosaic
of habitats (EU Habitats Directive) and importance for biodiversity,
including migratory bird species (part of the EU Birds Directive) and
intertidal vegetation (from mudflats to seagrass meadows and salt
marshes)..

@@

Before each flight, targets used as ground control points were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). Ground control points were used to
correct georeferencing imprecision of orthomosaics with an horizontal
and vertical accuracy of 10cm. A dGPS was also used to georeference
quadrats of 0.25 m², which assessed the presence or absence of five key
taxonomic classes of intertidal vegetation: Bacillariophyceae (benthic
diatoms forming biofilms at the sediment surface during low tide with
biofilm's size ranging from small patches (m²) to entire mudflats (km²);
henceforth: Benthic diatoms), Phaeophyceae (brown macroalgae generally
attached to rocks or other substrates able to form dense beds in the
intertidal zone; henceforth: Brown macroalgae), Magnoliopsida
(seagrasses, rooted flowering marine plants able to form extensive
meadows on soft sediments; henceforth: Seagrasses), Chlorophyceae (green
macroalgae, typically found attached to rocks or washed ashore;
henceforth: Green macroalgae), and Rhodophyceae (red macroalgae,
attached to hard substrates but can also be found on soft-bottom
substrate; henceforth: Red macroalgae). Only homogeneous vegetation
patches extending over several meters were selected as ground control
points. Pictures of each quadrat were uploaded online to the open-portal
Global Biodiversity Information Facility (GBIF) platform
\citep{BedeGbif}. Each photograph was also processed to estimate the
percent cover of each type of vegetation using an image processing
software \citep[ImageJ,][]{schneider2012nih}. Hyperspectral reflectance
signatures of each vegetation class were recorded using an ASD FieldSpec
HandHeld 2 spectroradiometer, which acquires reflectance between 325 and
1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures
served dual purposes: they validate the radiometric calibration of drone
data and contributed to misclassification reduction in photo
interpretations.

@@

\caption{\label{fig-vegetation}The five taxonomic classes of vegetation
used to train the Neural Network model and an example of their raw
spectral signatures at the spectral resolution of the Micasense RedEdge
Dual MX. A : Magnoliopsida (\emph{Zostera noltei}) ; B : Phaeophyceae
(\emph{Fucus sp.}) ; C : Rhodophyceae (\emph{Gracilaria
vermiculophylla}) ; D : Chlorophyceae (\emph{Ulva sp.}) ; E :
Bacillariophyceae (Benthic diatoms). Classes and species taxonomy
following the WORMS - World Register of Marine Species classfication.}

@@

The high-altitude flight over Gafanha covered a total area of
\textasciitilde1 km² (Figure~\ref{fig-GafHigh}). A channel contouring a
small island was masked in the prediction map. Most of the vegetation
area was classified as seagrass by the model, including patches with
brown leaves. Only a few pixels were classified as green macroalgae (F1
score of 0.55). Patches of red macroalgae were correctly classified (F1
score of 0.85). In the northern part of the site and near the land
edges, patches of the halophyte \emph{Sporobolus maritimus} (syn.
\emph{Spartina maritima)} were misclassified, either as seagrass or as
brown algae (F1 score of 0.77 and 0.71, respectively).

@@

\caption{\label{fig-Boat}RGB orthomosaic (Top) and Prediction (Bottom)
of the flight made in the inner part of Ria de Aveiro lagoon, Portugal.
The total extent of this flight was about 1.5 km² with a resolution of
80 mm per pixel. The zoom inserts cover an area equivalent to the size
of a 10-meter Sentinel-2 pixel.}

@@

The lowest performing site was Gafanha High (global accuracy of 75.45\%)
whereas Mataduços was the site with the most accurate prediction (global
accuracy of 98.05\%). Overall, the classes Phaeophyceae, Magnoliopsida,
Sediment and Rhodophyceae were correctly classified with a balanced
accuracy of 1, 0.96, 0.96 and 0.91 respectively. Bacillariophyceae was
the least accurate class (accuracy of 0.72 ) mainly due to confusion
with Magnoliopsida and Sediment.

@@

\subsection{Effect of the cover on the
prediction}\label{effect-of-the-cover-on-the-prediction}

Using the very high-resolution low-altitude flight (8 mm pixels), we
determined the minimal cover (\%) required to correctly classify a given
class within the corresponding high-altitude flight (8cm pixel
resolution ; Figure~\ref{fig-upscaling}).

@@

\caption{\label{fig-upscaling}Kernel density plot showing the proportion
of pixel well classified based on the percent cover of the class in high
altitude flight pixels of Gafanha, Portugal. Each subplot shows all the
pixels of the same classes on the high altitude flight. Cover (\%) of
classes was retrieved using the result of the classification of the low
altitude flight of Gafanha, Portugal.}

@@

A cover of at least 80 \% was sufficient to have all the 80 mm pixels
correctly classified, with the exception of Magnoliopsida which required
a higher cover (\textgreater90 \%) to be accurately classified.
Concerning the probability of each class, there is a linear relationship
between the percent cover and the confidence of the model to predict the
class. To predict green macroalgae with a model likelihood of 0.85, a
cover of 93 \% was needed, 90 \% for seagrass, 92 \% for red macroalgae
and 97 \% for benthic diatoms. When the vegetation cover of a given
class was 100 \%, coarser high-flight pixels were correctly classified
for all the classes except for Bare Sediment, which was only correctly
classified 80\% of the time. This phenomenon may be attributed to the
time gap between the two flights, allowing for microphytobenthos
migration to the sediment surface during low tide, consequently altering
the model's classification from bare sediment to Bacillariophyceae.

@@

The authors would like to thank the extensive revisions and comments
made by anonymous reviewers to improve the work. This work was supported
through the BiCOME (Biodiversity of the Coastal Ocean: Monitoring with
Earth Observation) project funded by the European Space Agency under the
`Earth Observation Science for Society' element of FutureEO-1
BIODIVERSITY+PRECURSORS call, contract No.~4000135756/21/I-EF. This work
was also supported by French Ministry of Research \& Higher Education
for funding first authors PhD. This study was partially funded by
FCT/MCTES (Fundação para a Ciência e a Tecnologia) under the project
UIDB/50017/2020 + UIDP/50017/2020 + LA/P/0094/2020 granted to CESAM and
through research contract CEECIND/00962/2017 (DOI:
10.54499/CEECIND/00962/2017/CP1459/CT0008) granted to AI Sousa. The work
was also partially funded by the project BIOPRADARIA
(MAR-01.04.02-FEAMP-0020), funded by the Operational Program MAR2020,
EMFF-European Maritime and Fisheries Fund, European Union, Portugal2020.
